---
title: "RMS Titanic"
author: "Sam Mendelson"
date: 2025-02-05
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- Class
- Sex
- Age
- Survived
- n

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Are there any differences?
  - There is a difference of 1 person: Wikipedia says that there were [1,317 passengers onboard](https://en.wikipedia.org/wiki/Titanic#Passengers) and [885 crew](https://en.wikipedia.org/wiki/Titanic#Crew), for a total of 2,202 souls.
- If yes, what might account for those differences?
  - Under both Crew and Passengers (in the Maiden Voyage section) Wikipedia uses the word "approximately" for the counts.
  - Passenger counts at the time varied between sources, too.

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
survivors <- df_titanic %>% filter(Survived == "Yes")

survivors %>%
  ggplot(aes(x = Class, y = n)) +
  geom_col(aes(fill = Sex), position = "dodge") +
  ggtitle("Titanic Survivors by Class and Sex")
```

**Observations**:

- It's clear that women were first in line for lifeboats, as far more women survived than men in all classes.
- The total number of survivors is highest in first class, followed by third class, then second class.
- Very few of the female crew survived. It is possible that this is due to the crew being made up of mostly men.
- The number of survivors in third class is higher than in second class, which is surprising given the common perception that third class passengers were the last to be saved.

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
    ggplot(aes(x = Class, y = Prop, fill = Sex)) +
    geom_col(position = "dodge") +
    ggtitle("Proportion of Titanic Survivors, by Class and Sex")
```

**Observations**:

- The plot shows that equal proportions of 1st and 2nd class men and women survived, which is incongruent with the historical record where women and children were loaded into lifeboats first.
- The proportion of female crew members who survived is higher than the proportion of 3rd class women who survived.
- Is there anything *fishy* going on in your plot?
  - One clear sign that something fishy is going on is that this chart appears to show that (at least in 1st and 2nd class), an equal proportion of men and women survived. However, we know this is untrue from historical records and our first chart.

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}

df_prop %>%
  filter(Survived == "Yes") %>%
    ggplot(aes(x = Class, y = Prop, Age)) +
    facet_grid(~Age) +
    geom_col(aes(fill = Sex), position = "dodge")
```

**Observations**:

- By splitting the chart by age, we see a plot that matches the historical record: women and children were loaded into lifeboats first.
- The proportion of survivors is highest in first class, followed by third class, then second class.
- It does show that 100% of children in 1st and 2nd class survived (which is not true), so there may be some data quality issues.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - Now, we see that tons and tons of men and women that survived in 1st and 2nd class were children, which were offloaded first (with adult women).
  - Since a much smaller percentage of 3rd class women survived, it's likely that they were carrying the 3rd class children, therefore the proportion of 3rd class children who survived is lower (which this graph supports).

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
