---
title: "Michelson Speed-of-light Measurements"
author: "Sam Mendelson"
date: 2025-02-13
output: 
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: When studying physical problems, there is an important
distinction between *error* and *uncertainty*. The primary purpose of
this challenge is to dip our toes into these factors by analyzing a real
dataset.

*Reading*: [Experimental Determination of the Velocity of
Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
(Optional)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to
measure the speed of light. His approach was a development upon the
method of Foucault[3], and resulted in a new estimate of
$v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very
close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this
challenge, you will analyze Michelson's original data, and explore some
of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code
chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the
value `LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition)
in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse()
```

*Data dictionary*:

-   `Date`: Date of measurement
-   `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair,
    1 = poor
-   `Temp`: Ambient temperature (Fahrenheit)
-   `Velocity`: Measured speed of light (km / s)

### **q1** Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n   | MeanVelocity |
|--------------|-----|--------------|
| 3            | 46  | 299860       |
| 2            | 39  | 299860       |
| 1            | 15  | 299810       |

```{r q1-task}
## TODO: Compute summaries
df_q1 <- df_michelson %>%
  group_by(Distinctness) %>%
  summarize(
    n = n(),
    MeanVelocity = mean(Velocity)
  )
df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**:

-   Interestingly, when distinctness is poor (1), the mean velocity is
    closest to the real (vacuum) speed of light, and as distinctness
    goes up, the mean velocity gets further from the real value.

-   It looks like Michelson's velocities were rounded to the nearest 10.

The `Velocity` values in the dataset are the speed of light *in air*;
Michelson introduced a couple of adjustments to estimate the speed of
light in a vacuum. In total, he added $+92$ km/s to his mean estimate
for `VelocityVacuum` (from Michelson (1880), pg. 141). While the
following isn't fully rigorous ($+92$ km/s is based on the mean
temperature), we'll simply apply this correction to all the observations
in the dataset.

### **q2** Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <- df_michelson %>%
  mutate(
    VelocityVacuum = Velocity + 92
  )

df_q2
```

As part of his study, Michelson assessed the various potential sources
of error, and provided his best-guess for the error in his
speed-of-light estimate. These values are provided in
`LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently,
Michelson believed the true value of the speed-of-light probably lay
between `LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and
`LIGHTSPEED_MICHELSON + LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

-   **Error** is the difference between a true value and an estimate of
    that value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
-   **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does
not know the error. The best they can do is quantify their degree of
uncertainty. We will learn some means of quantifying uncertainty in this
class, but for many real problems uncertainty includes some amount of
human judgment.[2]

### **q3** Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
lightspeed_error <- abs(LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON)
lightspeed_error_error <- abs(lightspeed_error - LIGHTSPEED_PM)
print(lightspeed_error)
print(lightspeed_error_error)
# compute the % error
print((abs(LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON)/LIGHTSPEED_MICHELSON*100))
```

**Observations**:

-   Michaelson's estimate of his error (51m/s) is about 1/3 of the real
    error (\~150m/s). However, in perspective of these numbers, his
    estimate of the speed of light is just 0.05% off.

The following plot shows all of Michelson's data as a [control
chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot
is common in manufacturing, where it is used to help determine if a
manufacturing process is under [statistical
control](https://en.wikipedia.org/wiki/Statistical_process_control).
Each dot is one of Michelson's measurements, and the grey line connects
the mean taken for each day. The same plot also shows simulated data
using a probability model. We'll get into statistics later in the
course; for now, let's focus on understanding what real and simulated
data tend to look like.

### **q4** Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**:

Similarities:

-   Both the simulated and real datasets stay within about 200 m/s of
    the real answer

-   The simulated data has much less variation, however its variation
    stays constant as time goes on

Differences:

-   In the real data, you can see Michelson dial in his estimates over
    time - estimates from early June vary far more than ones near the
    end of the month

-   Michelson's experiments are more grouped together by day - he
    clearly didn't have time to work on this every day

### **q5** You have access to a few other variables. Construct **at least three** visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?

```{r}
df_with_error <- 
  df_q2 %>%
  mutate(
    MichelsonError = abs(VelocityVacuum - LIGHTSPEED_VACUUM)
  )
df_with_error
```

#### Plot 1: Distinctness vs. Temperature

```{r}
df_with_error %>%
  ggplot(aes(Distinctness, Temp)) +
  geom_boxplot() + 
  ggtitle("Distinctness vs. Temperature")
```

**Observations**:

-   The median of the temperature increases as the distinctness
    increases
-   Since higher distinctness indicates a better estimate, one could
    conclude that the best estimates are taken around 80˚F - however,
    there are two big outliers around 60˚F

#### Plot 2: Distinctness vs. Accuracy

```{r}
df_with_error %>%
  ggplot(aes(Distinctness, Velocity)) + 
  geom_point() +
  annotate("text", x=1, y=LIGHTSPEED_MICHELSON + 15, label="Michelson's final estimate", size=5, color="red") +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted",
    color = "red"
  ) +
  annotate("text", x=2.5, y=LIGHTSPEED_MICHELSON - LIGHTSPEED_PM + 15, label="Lower error bound", size=5, color="blue") +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed",
    color="blue"
  ) +
  annotate("text", x=1.5, y=LIGHTSPEED_MICHELSON + LIGHTSPEED_PM + 15, label="Upper error bound", size=5, color="orange") +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed",
    color = "orange"
  ) +
  annotate("text", x=.7, y=LIGHTSPEED_VACUUM + 15, label="True velocity", size=5, color="purple") +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dashed",
    color = "purple"
  ) +
  ggtitle("Distinctness vs. Velocity") +
  ylab("Speed of light (measured, m/s)")
```

**Observations**:

-   Distinctness does not appear to correlate to Michelson's *idea* of
    accuracy, although it does seem to correlate to true accuracy.
-   For example, the most accurate (when accuracy means closest to the
    real estimate of the speed of light as measured in a vacuum)
    estimates were at distinctness 1, which Michelson perceived to be
    worst

#### Plot 3: Temperature vs. Accuracy

```{r}
df_with_error %>%
  ggplot(aes(Temp, MichelsonError, color = Distinctness)) + 
  geom_point() +
  ggtitle("Temperature vs. Michelson's Measurement Error (vs. vacuum)") +
  xlab("Temperature during measurement (˚F)") +
  ylab("Michelson's measurement error, compared to vacuum (m/s)")
```

**Observations**:

-   Neither temperature nor distinctiveness is not correlated in a
    relevant way to Michelson's error
-   It's important to keep in mind that this plot measures *error*
    only - the actual number is in the hundreds of thousands, therefore
    these errors are less than 0.5% off of the actual estimate

## Bibliography

-   [1] Michelson, [Experimental Determination of the Velocity of
    Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
    (1880) 
-   [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical
    Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf)
    (1986) 
-   [3] BYU video about a [Fizeau-Foucault
    apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to
    what Michelson used.
