---
title: "Gapminder"
author: "Sam Mendelson"
date: 2025-03-02
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)
library(ggbreak)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)

summary(gapminder)
```

**Observations**:

-   country
-   continent
-   year
-   lifeExp (life expectancy)
-   pop (population)
-   gdpPercap (GDP per capita)

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% pull(year) %>% max()
year_min <- gapminder %>% pull(year) %>% min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r}
# source: https://stackoverflow.com/a/33525389
is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

min_year_gapminder_with_outliers <- 
  gapminder %>%
  filter(year == year_min) %>%
  group_by(continent) %>%
  mutate(
    outlier = ifelse(is_outlier(gdpPercap), as.character(country), as.character(NA))
  )

min_year_gapminder_with_outliers %>%
  ggplot(aes(x = continent, y = gdpPercap)) + 
  scale_y_break(c(15000, 107000)) +
  geom_boxplot() +
  geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.1) +
  ggtitle("GDP Per Capita vs. Continent, 1952") +
  xlab("Continent") +
  ylab("GDP Per Capita (PPP$2017)")
```

**Observations**:

-   Countries with a higher GDP and a lower population appear as outliers
-   Some continents have far more variation in their GDP per capita than others (Europe, for example, has a higher variation than Oceania, for example).

**Difficulties & Approaches**:

-   Kuwait is a very powerful outlier, with their GDP around 108,000. To solve this, I added a break in the Y axis between 15,000 and 107,000.
-   Without labeling the outliers, another plot or would be needed. To solve that, I added labels for the outliers.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
# done by looking at the plot above: outliers are marked
min_year_gapminder_with_outliers %>%
  filter(outlier != is.na(outlier))
```

**Observations**:

-   Identify the outlier countries from q2
    -   Africa
        -   South Africa
        -   Gabon
        -   Angola
    -   Americas
        -   United States
        -   Canada
        -   Venezuela
    -   Asia
        -   Kuwait
        -   Bahrain
    -   Europe
        -   Switzerland

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

max_year_gapminder_with_outliers <-
  gapminder %>%
  filter(
    year == year_max
  ) %>%
  group_by(
    continent
  ) %>%
  mutate(
    outlier = ifelse(is_outlier(gdpPercap), as.character(country), as.character(NA))
  )

final_data <- rbind(min_year_gapminder_with_outliers, max_year_gapminder_with_outliers)

final_data %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  # scale_y_break(c(18000, 107000)) +
  geom_boxplot() +
  geom_point(
    data = filter(min_year_gapminder_with_outliers, !is.na(outlier)),
    mapping = aes(color = outlier),
    size = 2
  ) +
  geom_point(
    data = filter(max_year_gapminder_with_outliers, !is.na(outlier)),
    mapping = aes(color = outlier),
    size = 2
  ) +
  facet_wrap(vars(year), scales = "free") +
  xlab("Continent") +
  ylab("GDP Per Capita (PPP$2017)")
```

**Observations**:

-   There are far fewer outliers in the Americas in the 2007 data
-   GDP per capita values all over the world (except in Kuwait) have risen drastically

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  ggplot(aes(x = year, y = gdpPercap, group = country, color = continent)) + 
  geom_line() + 
  xlab("Year") +
  ylab("GDP Per Capita (PPP$2017)") +
  ggtitle("Year vs. GDP Per Capita, 1952-2007")
```

-   In general, GDP per capita rose over time
-   In general, African economies had lower growth than European economies

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  ggplot(aes(x = year, y = lifeExp)) + 
  geom_line(aes(group = country, color = continent)) +
  # geom_smooth(aes(group = continent, color = continent)) +
  xlab("Year") +
  ylab("Life Expectancy (years)") +
  ggtitle("Year vs. Life Expectancy By Country, 1952-2007")
```

-   Life expectancies tend to rise over time
-   Africa likely had a significant health impact in the 1990s as many African countries' life expectancies dropped

```{r q5-task3}
## TASK: Your third graph
gapminder %>%
  ggplot(aes(x = lifeExp, y = gdpPercap)) + 
  geom_point(aes(color = continent)) + 
  geom_smooth() +
  ylab("GDP Per Capita (PPP $2017)") +
  xlab("Life Expectancy (years)") +
  ggtitle("Life Expectancy vs. GDP Per Capita, 1952-2007")
```

-   In general, countries with higher GDP per capita have higher life expectancies, except for some outliers:
-   Some Asian outliers have extremely high GDP per capita values without absurdly high life expectancies
-   European and American countries with the highest GDP per capitas lead the chart for expectancies
