---
title: "Antibiotics"
author: "Sam Mendelson"
date: 2025-03-06
output:
  github_document
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Creating effective data visualizations is an *iterative*
process; very rarely will the first graph you make be the most
effective. The most effective thing you can do to be successful in this
iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely
dependent on *the question you are trying to answer*. A visual that is
totally ineffective for one question may be perfect for answering a
different question.

In this challenge, you will practice *iterating* on data visualization,
and will anchor the *assessment* of your visuals using two different
questions.

*Note*: Please complete your initial visual design **alone**. Work on
both of your graphs alone, and save a version to your repo *before*
coming together with your team. This way you can all bring a diversity
of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(ggrepel)
library(ggbreak)
```

*Background*: The data[1] we study in this challenge report the
[*minimum inhibitory
concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration)
(MIC) of three drugs for different bacteria. The smaller the MIC for a
given drug and bacteria pair, the more practical the drug is for
treating that particular bacteria. An MIC value of *at most* 0.1 is
considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin,
streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized
into a genus based on a number of features, including their resistance
to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()

## Pivot the data
df_pivoted <- df_antibiotics %>%
  pivot_longer(
    cols = c(penicillin, streptomycin, neomycin),
    names_to = "antibiotic"
  ) %>%
  group_by(antibiotic) %>%
  mutate(
    gram_symbol = if_else(gram == "positive", "+", "-"),
    value_norm = (value - min(value)) / (max(value) - min(value))
  ) %>%
  arrange(gram)

df_pivoted
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the
data** `df_antibiotics`. These **cannot** be simple variations on the
same graph; for instance, if two of your visuals could be made identical
by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16
bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of
ideas, and don't fixate on any one idea for too long. You will have a
chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be
helpful to `pivot_longer()` the data to make certain visuals easier to
construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. This means **it must be possible to identify each of the
16 bacteria by name.** You must also show whether or not each bacterium
is Gram positive or negative.

```{r q1.1, fig.width=10}
# WRITE YOUR CODE HERE
df_pivoted %>%
  # group_by(antibiotic) %>%
  ggplot(aes(x = fct_reorder(bacteria, value), y = value, color = gram, shape = antibiotic)) +
  geom_point(position = "dodge") +
  geom_hline(yintercept = 0.1) +
  # facet_wrap(~ antibiotic, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 270)
  ) +
  ylab("Minimum Inhibitory Concentration (MIC)") +
  xlab("Bacteria") +
  scale_y_log10() +
  ggtitle("Bacteria vs. Minimum Inhibitory Concentration (MIC) by Antibiotic")
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. This means **it must be possible to identify each of the
16 bacteria by name.** You must also show whether or not each bacterium
is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.2}
# WRITE YOUR CODE HERE
df_pivoted %>%
  ggplot(aes(x = fct_reorder(bacteria, value), y = antibiotic, fill = value_norm)) +
  geom_tile() +
  geom_text(aes(label = gram_symbol), color = "white", size = 4) +
  theme(
    axis.text.x = element_text(angle = 270)
  ) +
  scale_fill_gradient(name = "Normalized MIC") +
  xlab("Bacteria") +
  ylab("Antibiotic") +
  ggtitle("Normalized MIC by Bacteria and Antibiotic")
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.3}
# WRITE YOUR CODE HERE
df_pivoted %>%
  # filter(
  #   antibiotic == "penicillin"
  # ) %>%
  ggplot(aes(x = fct_reorder(bacteria, value), y = value_norm, color = antibiotic, shape = gram)) +
  theme(
    axis.text.x = element_text(angle = 270)
  ) +
  geom_point() +
  xlab("Bacteria") +
  ylab("Normalized MIC") +
  scale_y_log10() +
  ggtitle("Normalized MIC by Bacteria and Antibiotic")
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.4, fig.width = 10, fig.height = 10}
# WRITE YOUR CODE HERE
# lollipop showing MIC
df_pivoted %>%
  # filter(
  #   antibiotic == "penicillin"
  # ) %>%
  ggplot(aes(x = fct_reorder(bacteria, value), y = value)) +
  geom_col() +
  geom_hline(yintercept = 0.1) +
  facet_wrap(~ bacteria, scales = "free") +
  ylab("MIC") +
  scale_y_log10() +
  ggtitle("MIC by Bacteria and Antibiotic") +
  xlab("Bacteria")
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE
df_pivoted %>%
  # filter(
  #   antibiotic == "penicillin"
  # ) %>%
  ggplot(aes(x = fct_reorder(bacteria, value), y = value, color = gram)) +
  theme(
    axis.text.x = element_text(angle = 270)
  ) +
  geom_point() +
  geom_hline(yintercept = 0.1) +
  facet_wrap(~ gram, shrink = TRUE) +
  ylab("Normalized MIC") +
  ggtitle("Normalized MIC by Bacteria and Antibiotic, Separated by Gram") +
  xlab("Bacteria")
```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer
both Guiding Questions. Note that you must also identify which of your
five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in
answering the questions below. You may need to revise one or more of
your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most
effective at helping answer both guiding questions. **Use this as an
opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against
> bacteria of different genera and Gram stain?

*Observations*

\- What is your response to the question above?

Neomycin appears least effective against gram-positive strepto- and
diplococcus bacteria.

Penicillin appears most effective against gram-positive bacteria of all
different types.

Streptomycin appears particularly effective against gram-positive
staphyloccocus and bacilus bacteria.

-   Which of your visuals above (1 through 5) is **most effective** at
    helping to answer this question?

#1 or #2 have equal effectiveness here, although I think that #2 is
easier to answer.

\- Why?

#1 has an extra piece of information: exact MIC counts for each of the
bacterias. However, that information isn't needed for this question, so
looking for bright colors is far easier, and the fact that there is just
one plot in #2 rather than three (in #1) makes for an easier reading
experience.

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*,
and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis*
[2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus
> pneumoniae*?

*Observations*

\- What is your response to the question above?

I believe that a reason behind the renaming of Diplococcus pneumoniae
(D.p.) to Streptococcus pneumoniae (S.p.) is due to the fact that D.p.'s
responses to bacteria are very similar to other bacteria in the
streptococcus family, specifically Streptococcus viridans and
Streptococcus hemolyticus. It's also gram-positive, just like the rest
of the streptococcus family (at least the ones in the dataset).

-   Which of your visuals above (1 through 5) is **most effective** at
    helping to answer this question?

I used #2 to answer this question.

\- Why?

I really like how it shows almost all of the data in a single graph, and
I could quickly compare between the Streptococcus bacteria and the
Diplococcus pneumonia.

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide
antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009)
[link](https://www.americanscientist.org/article/thats-funny)
