---
title: "US Income"
author: "Sam Mendelson"
date: 2025-04-06
output:
  github_document:
    toc: true
prerequisites:
  - e-stat09-bootstrap
---

*Purpose*: We've been learning how to quantify uncertainty in estimates through the exercises; now its time to put those skills to use studying real data. In this challenge we'll use concepts like confidence intervals to help us make sense of census data.

*Reading*: - [Using ACS Estimates and Margin of Error](https://www.census.gov/data/academy/webinars/2020/calculating-margins-of-error-acs.html) (Optional, see the PDF on the page) - [Patterns and Causes of Uncertainty in the American Community Survey](https://www.sciencedirect.com/science/article/pii/S0143622813002518?casa_token=VddzQ1-spHMAAAAA:FTq92LXgiPVloJUVjnHs8Ma1HwvPigisAYtzfqaGbbRRwoknNq56Y2IzszmGgIGH4JAPzQN0) (Optional, particularly the *Uncertainty in surveys* section under the Introduction.)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

# Setup

<!-- ----------------------------------------------------------------------- -->

```{r setup}
library(tidyverse)
library(usmap)
library(usmapdata)
library(sf)
```

### **q1** Load the population data from c06; simply replace `filename_pop` below.

```{r q1-task}
## TODO: Give the filename for your copy of Table B01003
filename_pop <- "data/ACSDT5Y2023.B01003-Data.csv"

## NOTE: No need to edit
df_pop <-
  read_csv(
    filename_pop,
    skip = 1,
  ) %>% 
  rename(
    population_estimate = `Estimate!!Total`
  )

df_pop
```

You might wonder why the `Margin of Error` in the population estimates is listed as `*****`. From the [documentation (PDF link)](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&cad=rja&uact=8&ved=2ahUKEwj81Omy16TrAhXsguAKHTzKDQEQFjABegQIBxAB&url=https%3A%2F%2Fwww2.census.gov%2Fprograms-surveys%2Facs%2Ftech_docs%2Faccuracy%2FMultiyearACSAccuracyofData2018.pdf%3F&usg=AOvVaw2TOrVuBDlkDI2gde6ugce_) for the ACS:

> If the margin of error is displayed as ‘\*\*\*\*\*’ (five asterisks), the estimate has been controlled to be equal to a fixed value and so it has no sampling error. A standard error of zero should be used for these controlled estimates when completing calculations, such as those in the following section.

This means that for cases listed as `*****` the US Census Bureau recommends treating the margin of error (and thus standard error) as zero.

### **q2** Obtain median income data from the Census Bureau:

-   `Filter > Topics > Income and Poverty > Income and Poverty`
-   `Filter > Geography > County > All counties in United States`
-   Look for `Median Income in the Past 12 Months` (Table S1903)
-   Download the 2018 5-year ACS estimates; save to your `data` folder and add the filename below.

```{r q2-task}
## TODO: Give the filename for your copy of Table S1903
filename_income <- "data/ACSST5Y2018.S1903-Data.csv"

## NOTE: No need to edit
df_income <-
  read_csv(filename_income, skip = 1)
df_income
```

Use the following test to check that you downloaded the correct file:

```{r q2-tests}
## NOTE: No need to edit, use to check you got the right file.
assertthat::assert_that(
  df_income %>%
    filter(Geography == "0500000US01001") %>%
    pull(`Estimate!!Percent Distribution!!FAMILY INCOME BY FAMILY SIZE!!2-person families`)
  == 45.6
)

print("Well done!")
```

This dataset is in desperate need of some *tidying*. To simplify the task, we'll start by considering the `\\d-person families` columns first.

### **q3** Tidy the `df_income` dataset by completing the code below. Pivot and rename the columns to arrive at the column names `id, geographic_area_name, category, income_estimate, income_moe`.

*Hint*: You can do this in a single pivot using the `".value"` argument and a `names_pattern` using capture groups `"()"`. Remember that you can use an OR operator `|` in a regex to allow for multiple possibilities in a capture group, for example `"(Estimate|Margin of Error)"`.

```{r q3-task}
df_q3 <-
  df_income %>%
  select(
    Geography,
    contains("Geographic"),
    # This will select only the numeric d-person family columns;
    # it will ignore the annotation columns
    contains("median") & matches("\\d-person families") & !contains("Annotation of")
  ) %>%
  mutate(across(contains("median"), as.numeric)) %>%
  pivot_longer(
    cols = -c(Geography, `Geographic Area Name`),
    names_to = c(".value", "category"),
    names_pattern = "(Estimate|Margin of Error)!!Median income \\(dollars\\)(?: MOE)?!!FAMILY INCOME BY FAMILY SIZE!!(.*)"
  ) %>%
  rename(
    geographic_area_name = `Geographic Area Name`,
    income_estimate = Estimate,
    income_moe = `Margin of Error`
  ) %>%
## TODO: Pivot the data, rename the columns
  glimpse()
```

Use the following tests to check your work:

```{r q3-tests}
## NOTE: No need to edit
assertthat::assert_that(setequal(
  names(df_q3),
  c("Geography", "geographic_area_name", "category", "income_estimate", "income_moe")
))

assertthat::assert_that(
  df_q3 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_moe)
  == 6663
)

print("Nice!")
```

The data gives finite values for the Margin of Error, which is closely related to the Standard Error. The Census Bureau documentation gives the following relationship between Margin of Error and Standard Error:

$$\text{MOE} = 1.645 \times \text{SE}.$$

### **q4** Convert the margin of error to standard error. Additionally, compute a 99% confidence interval on income, and normalize the standard error to `income_CV = income_SE / income_estimate`. Provide these columns with the names `income_SE, income_lo, income_hi, income_CV`.

```{r q4-task}

df_q4 <- df_q3 %>%
  mutate(
    income_SE = income_moe / 1.645,
  ) %>%
  mutate(
    income_CV = income_SE / income_estimate,
    income_lo = income_estimate - 2.576 * income_SE,
    income_hi = income_estimate + 2.576 * income_SE
  )
```

Use the following tests to check your work:

```{r q4-tests}
## NOTE: No need to edit
assertthat::assert_that(setequal(
  names(df_q4),
  c("Geography", "geographic_area_name", "category", "income_estimate", "income_moe",
    "income_SE", "income_lo", "income_hi", "income_CV")
))

assertthat::assert_that(
  abs(
    df_q4 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_SE) - 4050.456
  ) / 4050.456 < 1e-3
)

assertthat::assert_that(
  abs(
    df_q4 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_lo) - 54513.72
  ) / 54513.72 < 1e-3
)

assertthat::assert_that(
  abs(
    df_q4 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_hi) - 75380.28
  ) / 75380.28 < 1e-3
)

assertthat::assert_that(
  abs(
    df_q4 %>%
    filter(Geography == "0500000US01001", category == "2-person families") %>%
    pull(income_CV) - 0.06236556
  ) / 0.06236556 < 1e-3
)

print("Nice!")
```

One last wrangling step: We need to join the two datasets so we can compare population with income.

### **q5** Join `df_q4` and `df_pop`.

```{r q5-task}
## TODO: Join df_q4 and df_pop by the appropriate column

df_data <- inner_join(df_q4, df_pop, by = "Geography")
```

# Analysis

<!-- ----------------------------------------------------------------------- -->

We now have both estimates and confidence intervals for `\\d-person families`. Now we can compare cases with quantified uncertainties: Let's practice!

### **q6** Study the following graph, making sure to note what you can *and can't* conclude based on the estimates and confidence intervals. Document your observations below and answer the questions.

```{r q6-task}
## NOTE: No need to edit; run and inspect
wid <- 0.5

df_data %>%
  filter(str_detect(geographic_area_name, "Massachusetts")) %>%
  mutate(
    county = str_remove(geographic_area_name, " County,.*$"),
    county = fct_reorder(county, income_estimate)
  ) %>%

  ggplot(aes(county, income_estimate, color = category)) +
  geom_errorbar(
    aes(ymin = income_lo, ymax = income_hi),
    position = position_dodge(width = wid)
  ) +
  geom_point(position = position_dodge(width = wid)) +

  coord_flip() +
  labs(
    x = "County",
    y = "Median Household Income"
  )
```

**Observations**:

-   Document your observations here.
    -   Income appears to coorrelate with family size - higher income -\> larger family
    -   However, larger families seem to also have larger confidence intervals, pointing to the fact that there are many reasons why families would grow
    -   Larger confidence intervals could point to higher income inequality - in Nantucket, Hampshire, and Dukes county, for example.
-   Can you confidently distinguish between household incomes in Suffolk county? Why or why not?
    -   Not particularly. The more people in a household, the larger the confidence interval is - while you might be able to identify a trend you can't confidently distinguish.
-   Which counties have the widest confidence intervals?
    -   Nantucket (Nantucket Island), Hampshire (UMass Amherst), Dukes (Martha's Vineyard), Berkshire (the Berkshires)

In the next task you'll investigate the relationship between population and uncertainty.

### **q7** Plot the standard error against population for all counties. Create a visual that effectively highlights the trends in the data. Answer the questions under *observations* below.

*Hint*: Remember that standard error is a function of *both* variability (e.g. variance) and sample size.

```{r q7-task}

df_data %>%
  group_by(category) %>%
  ggplot(aes(y = income_SE, x = population_estimate, color = category)) +
  scale_y_log10() +
  geom_point(alpha = 0.05) + 
  geom_smooth() + 
  xlab("Standard Error of Median Household Income") +
  ylab("Population Estimate (2018)")
```

**)Observations**:

-   What *overall* trend do you see between `SE` and population? Why might this trend exist?
    -   As population goes down, standard error goes down.
    -   This may exist because, when there are fewer people, there are fewer outliers (people with super high or super low incomes).
-   What does this *overall* trend tell you about the relative ease of studying small vs large counties?
    -   I glean that small and large(r) counties are different beasts, and that if a county has, for example, 1,000,000 people, depending on the goal of the analysis, you may need to break it up into meaningful chunks to be able to study it.

# Going Further

<!-- ----------------------------------------------------------------------- -->

Now it's your turn! You have income data for every county in the United States: Pose your own question and try to answer it with the data.

### **q8** Pose your own question about the data. Create a visualization (or table) here, and document your observations.

```{r q8-task, fig.retina = TRUE, fig.width = 12.5}
## TODO: Pose and answer your own question about the data
county_centroids <- usmapdata::centroid_labels(regions = "counties") %>%
  select("fips", "geom") %>%
  mutate(
    coords = st_coordinates(geom)
  ) %>%
  mutate(
    lon = coords[,"X"], # longitude / lon
    lat = coords[,"Y"] # latitude / lat
  ) %>%
  select(
    "fips",
    "geom",
    "lon",
    "lat"
  )

df_map <-
  df_data %>%
  mutate(fips = str_sub(Geography, -5)) %>%
  filter(category == "2-person families") %>%
  left_join(county_centroids, by = "fips") %>%
  distinct(fips, .keep_all = TRUE)

plot_usmap(
  data = df_map,
  values = "income_SE",
  regions = "counties"
  ) +
  scale_fill_gradient(
    trans = 'log',
    high = '#0072B2',     # change the colors as necessary
    low = 'white'
  ) +
  theme(legend.position = "right") +
  ggtitle("Median Income Estimate: Standard Error, 2018") +
  labs(fill = "Standard Error")
```

**Observations**:

-   This map looks at 2-person family standard error across the US (except for CT, which appears to be missing)
-   Standard error appears to be highest in counties with low populations, and appears highest in the west half of the country for some reason
-   I'd posit that this occurs due to those being some of the counties with the lowest populations

Ideas:

-   Compare trends across counties that are relevant to you; e.g. places you've lived, places you've been, places in the US that are interesting to you.
-   In q3 we tidied the median `\\d-person families` columns only.
    -   Tidy the other median columns to learn about other people groups.
    -   Tidy the percentage columns to learn about how many households of each category are in each county.
-   Your own idea!

# References

<!-- ----------------------------------------------------------------------- -->

[1] Spielman SE, Folch DC, Nagle NN (2014) Patterns and causes of uncertainty in the American Community Survey. Applied Geography 46: 147–157. <pmid:25404783> [link](https://www.sciencedirect.com/science/article/pii/S0143622813002518?casa_token=VddzQ1-spHMAAAAA:FTq92LXgiPVloJUVjnHs8Ma1HwvPigisAYtzfqaGbbRRwoknNqZ6Y2IzszmGgIGH4JAPzQN0)
