---
title: "Massachusetts Highway Stops"
author: "Sam Mendelson"
date: 2025-04-27
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup

<!-- -------------------------------------------------- -->

### **q1** Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "data/yg821jf8611_ma_statewide_2020_04_01.rds"
df_data <- readRDS(filename)
```

# EDA

<!-- -------------------------------------------------- -->

### **q2** Do your "first checks" on the dataset. What are the basic facts about this dataset?

```{r}
summary(df_data)
```

**Observations**:

-   What are the basic facts about this dataset?
    -   There are over 3.4 million observations - very cool
    -   The date range is between 1/1/07 and 12/31/15
    -   Subjects' ages range from 10-74 with an average of 36.47
    -   Most subjects are white, and there are over 2 men per woman
    -   There are then a number of boolean variables about the stop

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### **q3** Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?

```{r q3-task}
## TODO: Determine the factor levels for subject_race and raw_Race
df_races <- df_data %>%
  group_by(raw_Race, subject_race) %>%
  summarize(subject_race_count = n())
```

**Observations**:

-   What are the unique values for `subject_race`?
    -   asian/pacific islander
    -   black
    -   hispanic
    -   white
    -   other
    -   unknown
    -   NA
-   What are the unique values for `raw_Race`?
    -   A
    -   American Indian or Alaskan Native
    -   Asian or Pacific Islander
    -   Black
    -   Hispanic
    -   Middle Eastern or East Indian (South Asian)
    -   None - for no operator present citations
    -   White
-   What is the overlap between the two sets?
    -   Races that are present in `subject_race` are directly copied, however "other" and "unknown" comprise different groups.
    -   NAs are present in both.
-   What is the difference between the two sets?
    -   "A" and "American Indian or Alaskan Native" are marked as "other", and `None - for no operator present citations` are marked as "unknown" in subject_race.

### **q4** Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
## TODO: Devise your own way to test the hypothesis posed above.
df_data %>% filter(subject_race == "white" & raw_Race != "White")
```

**Observations**

Between the two hypotheses:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

-   `race_Raw` is most plausible as an unprocessed version of `subject_race`, as there are zero rows where `subject_race` == "white" and `raw_Race` != "White".
-   While this is not completely guaranteed, it would mean that across all 3.4+ million stops, ONLY white officers could have stopped white subjects, which is highly unlikely.

## Vis

<!-- ------------------------- -->

### **q5** Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)

**Observations**:

-   How does `arrest_rate` tend to vary with `subject_age`?

In general, arrest rates are highest for minors, however there are far fewer stops involving minors.

```{r}
df_data %>%
  group_by(subject_age) %>%
  summarize(arrest_rate = mean(arrest_made, na.rm = TRUE), num_obs = n()) %>%
  ggplot(aes(subject_age, arrest_rate, color = num_obs)) +
  geom_point() +
  ggtitle("Subject Age vs. Arrest Rate, MA State Police 2007-2015") +
  xlab("Subject Age") +
  ylab("Arrest Rate") +
  labs(color = "# Observations")
```

-   How does `arrest_rate` tend to vary with `subject_sex`?

It appears that arrest rates are highest for men.

```{r}
df_data %>%
  group_by(subject_sex) %>%
  summarize(arrest_rate = mean(arrest_made, na.rm = TRUE), num_obs = n()) %>%
  ggplot(aes(subject_sex, arrest_rate)) +
  geom_col() +
  ggtitle("Subject Sex vs. Arrest Rate, MA State Police 2007-2015") +
  xlab("Subject Sex") +
  ylab("Arrest Rate") +
  labs(color = "# Observations")
```

-   How does `arrest_rate` tend to vary with `subject_race`?

Note that for clarity, some outliers have been removed.

Overall, arrest rates are highest for the hispanic race, then black, then white.

```{r}
df_data %>%
  group_by(subject_race) %>%
  summarize(arrest_rate = mean(arrest_made, na.rm = TRUE), num_obs = n()) %>%
  ggplot(aes(subject_race, arrest_rate)) +
  geom_col() +
  ggtitle("Subject Sex vs. Arrest Rate, MA State Police 2007-2015") +
  xlab("Subject Sex") +
  ylab("Arrest Rate")
```

# Modeling

<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### **q6** Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()
```

**Observations**:

-   Which `subject_race` levels are included in fitting the model?
    -   Only white, black, and hispanic.
-   Which `subject_race` levels have terms in the model?
    -   Hispanic and white.

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### **q7** Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race
fit_q7 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ) %>%
      mutate(factor = fct_relevel(subject_race, "white")),
    family = "binomial"
  )

fit_q7 %>% tidy()
```

**Observations**:

-   Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
    -   According to this model, hispanic people have the highest probability of being arrested, and white people have the lowest probability.
-   What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
    -   There could be less hispanic people in Massachusetts, therefore they would be less likely to have many stops that do not result in an arrest (like speeding/parking tickets).
    -   Hispanic people could be less likely to drive, therefore also resulting in less stops that do NOT result in an arrest.
    -   There could be bias from the arresting officers, whose biases could result in more hispanic people getting arrested.
-   Look at the set of variables in the dataset; do any of the columns relate to a potential explanation you listed?
    -   There are some contraband columns, relating to types of contraband that officers may have found. It's possible that biases would result in the fact that more arrests happen after a contraband discovery from a hispanic subject.
    -   Reason for stop could be important, but this would require much more work to categorize (it's comma-separated)

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### **q8** Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q8 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ) %>%
      mutate(factor = fct_relevel(subject_race, "white")),
    family = "binomial"
  )

fit_q8 %>% tidy()
```

**Observations**:

-   How does controlling for found contraband affect the `subject_race` terms in the model?
    -   The presence of contraband is now the largest factor, which has diminished the effect of the race term significantly.
-   What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
    -   The finding of contraband tells us that the stop was likely to result in an arrest.
    -   It does NOT tell us what the contraband was, how much there was, or any other factors that could contribute to an arrest.

### **q9** Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

How do frisks influence arrests?

```{r q9-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q9 <-
  glm(
    formula = arrest_made ~ frisk_performed + search_conducted,
    data = df_data %>%
      mutate(factor = fct_relevel(subject_race, "white")) %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q9 %>% tidy()
```

**Observations**:

-   My question is: "How do the odds of a frisk or a search influence the odds of an arrest?"
-   It looks like the odds of a **frisk** decreases the chance of arrest, but the odds of a **search** increase the odds.
-   I was really curious why this was, so I did some research on the legal difference between a search and frisk.
    -   The difference between a frisk and a search is that the purpose of a frisk is limited to finding weapons hidden on the suspect to ensure the personal safety of police officers, its scope is limited to weapons and not for the discovery of other evidence; a search generally requires a search warrant , and its purpose is mainly to find evidence of illegal criminal activity. ([source](https://www.law.cornell.edu/wex/frisk#:~:text=The%20difference%20between%20a%20frisk,warrant%20%2C%20and%20its%20purpose%20is))
-   Since a search requires a warrant, it makes sense that the arrest probability would rise: if a judge believes that there may be evidence for an arrest, then there is a reasonable chance that one is made after the search.
-   And, since a frisk helps to ensure that there are no weapons on a subject (or that those weapons are disclosed), officers might feel safer and therefore less likely to arrest the subject.
-   Not really an observation, but it was interesting to learn the (legal) difference between a search and a frisk.

## Further Reading

<!-- -------------------------------------------------- -->

-   Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
